---
title: |
  | "PH700A Week 14"
subtitle: |
  | Lab Exercise 14
  | PH 700A, Spring 2025
author: "rycalvo"
format: html
editor: source
---

## Assignment Background

To create, validate, and test a model to accurately predict admission to the hospital based on symptoms and factors collected in the Emergency Room.

_Research Question_: Among patients transported to the emergency department via ambulance, what are the best set of predictors that can be used to determine admission to the hospital.

## Data Loading
```{r}
#| label: Data Loading from Github
#| include: false

library(data.table)
library(tidyverse)
library(explore)

df <- fread("https://github.com/core-academics/sdsu.ph700a/raw/main/data/mimic4ed/week14_df.csv.gz")

# Removes the one column from the export
df$V1 <- NULL

# catvars gender:alzheimer, eval:renaluri, pain_cat
df <- df %>% mutate_at(vars(-temperature, -heartrate, -resprate, -o2sat, -sbp, -dbp, -edloshrs), as.factor)

# Remove columns that do not vary
df <- df %>%
  select_if(~ !all(. == first(.)))
```

## Subsetting Data
```{r}
#| include: true
#| label: data subsetting

library(rsample)

set.seed(20250430)

var_subselection <- df %>%
  select(stay_id, gender, disposition, alzheimer, trauma, visual, edloshrs, temperature:acuity, cardiac)

subsets.tagged <- initial_validation_split(var_subselection, prop = c(0.5, 0.25))
  
train <- training(subsets.tagged)
valid <- validation(subsets.tagged)
test <- testing(subsets.tagged)
```

The primary outcome is `disposition`. The `stay_id` variable uniquely identifies rows.

## Objective

Please evaluate any combination of predictors using any modeling engine (logistic regression is the default) to identify the best set of predictors.

Please assess calibration and model performance in your code.